---
title: "Quantified Self Movement Analysis"
date: "14/1/2021"
output:
  html_document: default
  #md_document: default
---

This document contents the next items to be evaluated:

  1. Code for reading in the dataset and/or processing the data
  2. Histogram of the total number of steps taken each day
  3. Mean and median number of steps taken each day
  4. Time series plot of the average number of steps taken
  5. The 5-minute interval that, on average, contains the maximum number of steps
  6. Code to describe and show a strategy for imputing missing data
  7. Histogram of the total number of steps taken each day after missing values are imputed
  8. Panel plot comparing the average number of steps taken per 5-minute interval across     weekdays and weekends
  9.  All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

Loading the libraries to handle the data

```{r, echo=TRUE, results='asis'}
library(dplyr)
library(ggplot2)
```

1. Loading and preprocessing the data

    ```{r setup, echo=TRUE}
    activity<- read.csv("activity.csv", header = T)
    ```

2. What is mean total number of steps taken per day?


    ```{r, echo=TRUE}
    activity1<- activity %>%
      group_by(date) %>%
      summarise(TSteps = sum(steps))
    
    hist(activity1$TSteps,
         main = "Total Steps per Day",
         xlab = "Number of Steps")
    ```
     
3. Calculate and report the mean and median of the total number of steps taken per day.

    ```{r, echo=TRUE}
    activity2<- activity %>%
      group_by(date) %>%
      summarise(Mean = mean(steps, na.rm=T),
                Median = median(steps, na.rm = T))%>%
      print()
    ```
  
4. What is the average daily activity pattern?

    ```{r, echo=TRUE}
    activity3<- activity %>%
      group_by(interval) %>%
      summarise(Mean = mean(steps, na.rm = TRUE))
    
    plot(activity3$interval, activity3$Mean, type="l", xlab="5-minute Interval", ylab="Average Number Steps")
    
    ```

5. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

    ```{r, echo=TRUE}
    activity3[which.max(activity3$Mean),]
    ```

6. Code to describe and show a strategy for imputing missing data

    This code replaces the missing values for the calculated mean by interval obtained in the point 4:
    
    ```{r, echo=TRUE}
    missingValues <- is.na(activity$steps)
    activity4 <- transform(activity,
                              steps = ifelse(is.na(activity$steps),
                                             activity3$Mean[match(activity$interval, 
                                                                        activity3$interval)],
                                             activity$steps))
    ```

7. Histogram of the total number of steps taken each day after missing values are imputed

    ```{r, echo=TRUE}
    activity5<- activity4 %>%
      group_by(date) %>%
      summarise(TSteps = sum(steps))
    
    hist(activity5$TSteps,
         main = "Total Steps per Day",
         xlab = "Number of Steps")
    ```
8. Panel plot comparing the average number of steps taken per 5-minute interval across     weekdays and weekends

    Factoring the data in weekdays and weekends with the function below:

    ```{r, echo=TRUE}
    Day <- function(date) {
      day <- weekdays(date)
      if (day %in% c('lunes', 'martes', 'miércoles', 'jueves', 'viernes'))
        return ("weekeday")
      else if (day %in% c('sábado', 'domingo'))
        return ("weekend")
      else
        stop ("Invalid Date Format.")
    }
    ```

    Applying the function to the data creating a new column "day":
    
    ```{r, echo=TRUE}
    activity4$date <- as.Date(activity4$date)
    activity4$day <- sapply(activity4$date, FUN = Day)
    ```
    
    Plotting the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis) for the imputing data. 

    ```{r, echo=TRUE}
    activity6<- activity4 %>%
      group_by(day, interval) %>%
      summarise(Mean = mean(steps, na.rm = TRUE))
    
    ggplot(activity6, aes(interval, Mean)) + 
      geom_line() +
      facet_grid(day ~ .) +
      ggtitle("Average Daily Activity Pattern") +
      xlab("5-minute Interval") +
      ylab("Average Number of Steps") +
      theme(plot.title = element_text(hjust = 0.5))
    ```
